---
title: "How to build a parsnip model"
categories:
  - developer tools
type: learn-subsection
weight: 2
description: | 
  Create a parsnip model function from an existing model implementation.
toc: true
toc-depth: 2
include-after-body: ../../../resources.html
---

```{r}
#| label: "setup"
#| include: false
#| message: false
#| warning: false
rm(list = ls())
setwd("~/Documents/R_Packages/sandbox/bcn-tests/tidymodels.org/learn/develop/models")
source(here::here("common.R"))
```

```{r eval=FALSE}
# Enable repository from techtonique
options(repos = c(
  techtonique = 'https://techtonique.r-universe.dev',
  CRAN = 'https://cloud.r-project.org'))
  
# Download and install bcn in R
install.packages('bcn')
```

```{r}
#| label: "load"
#| include: false
#| message: false
#| warning: false
library(bcn)
library(tidymodels)

pkgs <- c("tidymodels", "bcn", "modeldata")

theme_set(theme_bw() + theme(legend.position = "top"))
```

## Introduction

`r article_req_pkgs(pkgs)`

The parsnip package constructs models and predictions by representing those actions in expressions. There are a few reasons for this:

-   It eliminates a lot of duplicate code.
-   Since the expressions are not evaluated until fitting, it eliminates many package dependencies.

A parsnip model function is itself very general. For example, the `logistic_reg()` function itself doesn't have any model code within it. Instead, each model function is associated with one or more computational *engines*. These might be different R packages or some function in another language (that can be evaluated by R).

This article describes the process of creating a new model function. Read [guidelines on creating modeling packages](https://tidymodels.github.io/model-implementation-principles/) to understand the general themes and conventions that are used.

## An example model

As an example, we'll create a function for *Boosted Configuration Networks*. We'll focus on `bcn::bcn`:

```{r}
#| label: "bcn-str"
str(bcn::bcn)
```

We'll name our function `bcn_model`.

## Aspects of models

Before proceeding, it helps to review how parsnip categorizes models:

-   The model *type* is related to the structural aspect of the model. For example, the model type `linear_reg` represents linear models (slopes and intercepts) that model a numeric outcome. Other model types in the package are `nearest_neighbor`, `decision_tree`, and so on.

-   Within a model type is the *mode*, related to the modeling goal. Currently the two modes in the package are regression and classification. Some models have methods for both models (e.g. nearest neighbors) while others have only a single mode (e.g. logistic regression).

-   The computation *engine* is a combination of the estimation method and the implementation. For example, for linear regression, one engine is `"lm"` which uses ordinary least squares analysis via the `lm()` function. Another engine is `"stan"` which uses the Stan infrastructure to estimate parameters using Bayes rule.

**When adding a model into parsnip, the user has to specify which modes and engines are used**. The package also enables users to add a new mode or engine to an existing model.

## The general process

The parsnip package stores information about the models in an internal environment object. The environment can be accessed via the function `get_model_env()`. The package includes a variety of functions that can get or set the different aspects of the models.

If you are adding a new model from your own package, you can use these functions to add new entries into the model environment.

### Step 1. Register the model, modes, and arguments

We will add the bcn model using the model type `bcn_model`. Since this is a classification method, we only have to register a single mode:

```{r}
#| label: "bcn-clf"
library(tidymodels)
try(set_new_model("bcn_model"), silent = TRUE)
set_model_mode(model = "bcn_model", mode = "classification")
set_model_engine(
  "bcn_model", 
  mode = "classification", 
  eng = "bcn"
)
set_dependency("bcn_model", eng = "bcn", pkg = "bcn")
```

These functions should silently finish. There is also a function that can be used to show what aspects of the model have been added to parsnip:

```{r}
#| label: "bcn-show-1"
show_model_info("bcn_model")
```

The next step would be to declare the main arguments to the model. These are declared independent of the mode. To specify the argument, there are a few slots to fill in:

-   The name that parsnip uses for the argument. In general, we try to use non-jargony names for arguments (e.g. "penalty" instead of "lambda" for regularized regression). We recommend consulting [the model argument table available here](/find/parsnip/) to see if an existing argument name can be used before creating a new one.

-   The argument name that is used by the underlying modeling function.

-   A function reference for a *constructor* that will be used to generate tuning parameter values. This should be a character vector with a named element called `fun` that is the constructor function. There is an optional element `pkg` that can be used to call the function using its namespace. If referencing functions from the dials package, quantitative parameters can have additional arguments in the list for `trans` and `range` while qualitative parameters can pass `values` via this list.

-   A logical value for whether the argument can be used to generate multiple predictions for a single R object. For example, for boosted trees, if a model is fit with 10 boosting iterations, many modeling packages allow the model object to make predictions for any iterations less than the one used to fit the model. In general this is not the case so one would use `has_submodels = FALSE`.

For `bcn::bcn()`, the main tuning parameters are:

-   \`B\`: a numeric, the number of iterations of the algorithm

-   \`nu\`: a numeric, the learning rate of the algorithm

-   \`col_sample\`: a numeric in \[0, 1\], the percentage of columns adjusted at each iteration

-   \`lam\`: a numeric, defining lower and upper bounds for neural network's weights

-   \`r\`: a numeric, with 0 \< r \< 1. Controls the convergence rate of residuals

-   \`tol\`: a numeric, convergence tolerance for an early stopping

Which we will rewrite as.

```{r}
#| label: "bcn-args"
parsnip::set_model_arg(
  model = "bcn_model",
  eng = "bcn",
  parsnip = "",
  original = "B",
  func = list(pkg = "bcn", fun = "bcn"),
  has_submodel = FALSE
)

parsnip::set_model_arg(
  model = "bcn_model",
  eng = "bcn",
  parsnip = "learn_rate",
  original = "nu",
  func = list(pkg = "bcn", fun = "bcn"),
  has_submodel = FALSE
)

parsnip::set_model_arg(
  model = "bcn_model",
  eng = "bcn",
  parsnip = "",
  original = "col_sample",
  func = list(pkg = "bcn", fun = "bcn"),
  has_submodel = FALSE
)

parsnip::set_model_arg(
  model = "bcn_model",
  eng = "bcn",
  parsnip = "",
  original = "lam",
  func = list(pkg = "bcn", fun = "bcn"),
  has_submodel = FALSE
)

parsnip::set_model_arg(
  model = "bcn_model",
  eng = "bcn",
  parsnip = "",
  original = "r",
  func = list(pkg = "bcn", fun = "bcn"),
  has_submodel = FALSE
)

parsnip::set_model_arg(
  model = "bcn_model",
  eng = "bcn",
  parsnip = "",
  original = "tol",
  func = list(pkg = "bcn", fun = "bcn"),
  has_submodel = FALSE
)

show_model_info("bcn_model")
```

### Step 2. Create the model function

This is a fairly simple function that can follow a basic template. The main arguments to our function will be:

-   The mode. If the model can do more than one mode, you might default this to "unknown". In our case, since it is only a classification model, it makes sense to default it to that mode so that the users won't have to specify it.

A basic version of the function is:

```{r}
#| label: "model-fun"
bcn_model <- function(mode = "classification", B = 10, nu = 0.1, 
                      col_sample = 1, lam = 0.1, r = 0.3, tol = 0) {
    
    # Check for correct mode
    if (mode  != "classification") {
      rlang::abort("`mode` should be 'classification'")
    }
    
    # Capture the arguments in quosures
    args <- list(B = rlang::enquo(B),
                 nu = rlang::enquo(nu),
                 col_sample = rlang::enquo(col_sample),
                 lam = rlang::enquo(lam),
                 r = rlang::enquo(r),
                 tol = rlang::enquo(tol))
    
    # Save some empty slots for future parts of the specification
    new_model_spec(
      "bcn_model",
      args = args,
      eng_args = NULL,
      mode = mode,
      method = NULL,
      engine = NULL
    )
  }
```

This is pretty simple since the data are not exposed to this function.

::: callout-warning
We strongly suggest favoring `rlang::abort()` and `rlang::warn()` over `stop()` and `warning()`. The former return better traceback results and have safer defaults for handling call objects.
:::

### Step 3. Add a fit module

Now that parsnip knows about the model, mode, and engine, we can give it the information on fitting the model for our engine. The information needed to fit the model is contained in another list. The elements are:

-   `interface` is a single character value that could be "formula", "data.frame", or "matrix". This defines the type of interface used by the underlying fit function (`bcn::bcn`, in this case). This helps the translation of the data to be in an appropriate format for the that function.

-   `protect` is an optional list of function arguments that **should not be changeable** by the user. In this case, we probably don't want users to pass data values to these arguments (until the `fit()` function is called).

-   `func` is the package and name of the function that will be called. If you are using a locally defined function, only `fun` is required.

-   `defaults` is an optional list of arguments to the fit function that the user can change, but whose defaults can be set here. This isn't needed in this case, but is described later in this document.

For the first engine:

```{r}
#| label: "fit-mod"
try(set_fit(
  model = "bcn_model",
  eng = "bcn",
  mode = "classification",
  value = list(
    interface = "matrix",
    func = c(pkg = "bcn", fun = "bcn"),
    defaults = list(B = 10, nu = 0.1, 
                      col_sample = 1, lam = 0.1, r = 0.3, tol = 0),
    #protect = list()
  )
), silent = FALSE)

show_model_info("bcn_model")
```

We also set up the information on how the predictors should be handled. These options ensure that the data that parsnip gives to the underlying model allows for a model fit that is as similar as possible to what it would have produced directly.

-   `predictor_indicators` describes whether and how to create indicator/dummy variables from factor predictors. There are three options: `"none"` (do not expand factor predictors), `"traditional"` (apply the standard `model.matrix()` encodings), and `"one_hot"` (create the complete set including the baseline level for all factors).

-   `compute_intercept` controls whether `model.matrix()` should include the intercept in its formula. This affects more than the inclusion of an intercept column. With an intercept, `model.matrix()` computes dummy variables for all but one factor level. Without an intercept, `model.matrix()` computes a full set of indicators for the first factor variable, but an incomplete set for the remainder.

-   `remove_intercept` removes the intercept column *after* `model.matrix()` is finished. This can be useful if the model function (e.g. `lm()`) automatically generates an intercept.

-   `allow_sparse_x` specifies whether the model can accommodate a sparse representation for predictors during fitting and tuning.

```{r}
set_encoding(
  model = "bcn_model",
  mode = "classification",
  eng = "bcn",
  options = list(predictor_indicators = "none", 
                 compute_intercept = FALSE,
                 remove_intercept = FALSE,
                 allow_sparse_x = FALSE)
)
```

### Step 4. Add modules for prediction

Similar to the fitting module, we specify the code for making different types of predictions. To make hard class predictions, the `class` object contains the details. The elements of the list are:

-   `pre` and `post` are optional functions that can preprocess the data being fed to the prediction code and to postprocess the raw output of the predictions. These won't be needed for this example, but a section below has examples of how these can be used when the model code is not easy to use. If the data being predicted has a simple type requirement, you can avoid using a `pre` function with the `args` below.
-   `func` is the prediction function (in the same format as above). In many cases, packages have a predict method for their model's class but this is typically not exported. In this case (and the example below), it is simple enough to make a generic call to `predict()` with no associated package.
-   `args` is a list of arguments to pass to the prediction function. These will most likely be wrapped in `rlang::expr()` so that they are not evaluated when defining the method. For bcn, the code would be `predict(object, newdata, type = "class")`. What is actually given to the function is the parsnip model fit object, which includes a sub-object called `fit()` that houses the bcn model object. If the data need to be a matrix or data frame, you could also use `newdata = quote(as.data.frame(newdata))` or similar.

The parsnip prediction code will expect the result to be an unnamed character string or factor. This will be coerced to a factor with the same levels as the original data.

To add this method to the model environment, a similar `set()` function is used:

```{r}
#| label: "mds-class"
class_info <- list(
    pre = NULL,
    post = NULL,
    func = c(pkg = "bcn", fun = "predict"),
    args =
      # These lists should be of the form:
      # {predict.bcn argument name} = {values provided from parsnip objects}
      list(
        # We don't want the first two arguments evaluated right now
        # since they don't exist yet. `type` is a simple object that
        # doesn't need to have its evaluation deferred. 
        object = quote(object),
        newdata = quote(newx),
        type = "class"
      )
  )

set_pred(
  model = "bcn_model",
  mode = "classification",
  eng = "bcn",
  type = "class",
  value = class_info
)

show_model_info("bcn_model")
```

A similar call can be used to define the class probability module (if they can be computed). The format is identical to the `class` module but the output is expected to be a tibble with columns for each factor level.

As an example of the `post` function, the data frame created by `bcn:::predict.bcn()` will be converted to a tibble. The arguments are `x` (the raw results coming from the predict method) and `object` (the parsnip model fit object). The latter has a sub-object called `lvl` which is a character string of the outcome's factor levels (if any).

We register the probability module. There is a template function that makes this slightly easier to format the objects:

```{r}
#| label: "bcn-prob"
prob_info <- pred_value_template(
    pre = NULL,
    post = NULL,
    func = c(pkg = "bcn", fun = "predict"),
    type = "prob"
  )

set_pred(
  model = "bcn_model",
  eng = "bcn",
  mode = "classification",
  type = "prob",
  value = prob_info
)

show_model_info("bcn_model")
```

If this model could be used for regression situations, we could also add a "numeric" module. For `pred`, the model requires an unnamed numeric vector output (usually).

Examples are [here](https://github.com/tidymodels/parsnip/blob/master/R/linear_reg_data.R) and [here](https://github.com/tidymodels/parsnip/blob/master/R/rand_forest_data.R).

### Does it work?

As a developer, one thing that may come in handy is the `translate()` function. This will tell you what the model's eventual syntax will be.

For example:

```{r}
#| label: "bcn-code"

B <- 5L

bcn_model(B = B) %>% translate(engine = "bcn")
```

Let's try it on a data set from the modeldata package:

```{r}
#| label: "bcn-data"
penguins_ <- as.data.frame(palmerpenguins::penguins)

replacement <- median(palmerpenguins::penguins$bill_length_mm, na.rm = TRUE)
penguins_$bill_length_mm[is.na(palmerpenguins::penguins$bill_length_mm)] <- replacement

replacement <- median(palmerpenguins::penguins$bill_depth_mm, na.rm = TRUE)
penguins_$bill_depth_mm[is.na(palmerpenguins::penguins$bill_depth_mm)] <- replacement

replacement <- median(palmerpenguins::penguins$flipper_length_mm, na.rm = TRUE)
penguins_$flipper_length_mm[is.na(palmerpenguins::penguins$flipper_length_mm)] <- replacement

replacement <- median(palmerpenguins::penguins$body_mass_g, na.rm = TRUE)
penguins_$body_mass_g[is.na(palmerpenguins::penguins$body_mass_g)] <- replacement

# replacing NA's by the most frequent occurence
penguins_$sex[is.na(palmerpenguins::penguins$sex)] <- "male" # most frequent

print(summary(penguins_))
print(sum(is.na(penguins_)))

# one-hot encoding
penguins_mat <- model.matrix(species ~., data=penguins_)[,-1]
penguins_mat <- cbind(penguins_$species, penguins_mat)
penguins_mat <- as.data.frame(penguins_mat)
colnames(penguins_mat)[1] <- "species"

print(head(penguins_mat))
print(tail(penguins_mat))
```

```{r}
#| label: "fit-bcn"

y <- as.integer(penguins_mat$species)
X <- as.matrix(penguins_mat[,2:ncol(penguins_mat)])

n <- nrow(X)
p <- ncol(X)


set.seed(1234)
index_train <- sample(1:n, size=floor(0.8*n))
X_train <- X[index_train, ]
y_train <- factor(y[index_train])
X_test <- X[-index_train, ]
y_test <- factor(y[-index_train])

bcn_spec <- bcn_model(B = B) %>% set_engine("bcn")
bcn_spec

bcn_fit <- bcn_spec %>% fit_xy(x = X_train, y = y_train, data = example_train, engine = "bcn")
#bcn_fit

#predict(bcn_fit, new_data = example_test, type = "prob") %>%
#  bind_cols(example_test %>% select(Class))

#predict(bcn_fit, new_data = example_test) %>% 
# bind_cols(example_test %>% select(Class))
```

```{}
```

## Your model, tuning parameters, and you

The tune package can be used to find reasonable values of model arguments via tuning. There are some S3 methods that are useful to define for your model. `bcn_model()` has one main tuning parameter: `sub_classes`. To work with tune it is *helpful* (but not required) to use an S3 method called `tunable()` to define which arguments should be tuned and how values of those arguments should be generated.

`tunable()` takes the model specification as its argument and returns a tibble with columns:

-   `name`: The name of the argument.

-   `call_info`: A list that describes how to call a function that returns a dials parameter object.

-   `source`: A character string that indicates where the tuning value comes from (i.e., a model, a recipe etc.). Here, it is just `"model_spec"`.

-   `component`: A character string with more information about the source. For models, this is just the name of the function (e.g. `"bcn_model"`).

-   `component_id`: A character string to indicate where a unique identifier is for the object. For a model, this is indicates the type of model argument (e.g. "main").

The main piece of information that requires some detail is `call_info`. This is a list column in the tibble. Each element of the list is a list that describes the package and function that can be used to create a dials parameter object.

For example, for a nearest-neighbors `neighbors` parameter, this value is just:

```{r}
#| label: "mtry"
info <- list(pkg = "dials", fun = "neighbors")

# FYI: how it is used under-the-hood: 
new_param_call <- rlang::call2(.fn = info$fun, .ns = info$pkg)
rlang::eval_tidy(new_param_call)
```

For `bcn_model()`, a dials object is needed that returns an integer that is the number of sub-classes that should be create. We can create a dials parameter function for this:

```{r}
#| label: "sub-classes"
sub_classes <- function(range = c(1L, 10L), trans = NULL) {
  new_quant_param(
    type = "integer",
    range = range,
    inclusive = c(TRUE, TRUE),
    trans = trans,
    label = c(sub_classes = "# Sub-Classes"),
    finalize = NULL
  )
}
```

If this were in the dials package, we could use:

```{r}
#| label: "tunable"
tunable.bcn_model <- function(x, ...) {
  tibble::tibble(
    name = c("sub_classes"),
    call_info = list(list(pkg = NULL, fun = "sub_classes")),
    source = "model_spec",
    component = "bcn_model",
    component_id = "main"
  )
}
```

Once this method is in place, the tuning functions can be used:

```{r}
#| label: "tune-bcn"
#| message: false
bcn_spec <- 
  bcn_model(sub_classes = tune()) %>% 
  set_engine("bcn")

set.seed(452)
cv <- vfold_cv(example_train)
bcn_tune_res <- bcn_spec %>%
  tune_grid(Class ~ ., cv, grid = 4)
show_best(bcn_tune_res, metric = "roc_auc")
```

## Pro-tips, what-ifs, exceptions, FAQ, and minutiae

There are various things that came to mind while developing this resource.

**Do I have to return a simple vector for `predict` and `predict_class`?**

Previously, when discussing the `pred` information:

> For `pred`, the model requires an unnamed numeric vector output **(usually)**.

There are some models (e.g. `glmnet`, `plsr`, `Cubist`, etc.) that can make predictions for different models from the same fitted model object. We want to facilitate that here so, for these cases, the current convention is to return a tibble with the prediction in a column called `values` and have extra columns for any parameters that define the different sub-models.

For example, if I fit a linear regression model via `glmnet` and get four values of the regularization parameter (`lambda`):

```{r}
#| label: "glmnet"
#| eval: false
linear_reg() %>%
  set_engine("glmnet", nlambda = 4) %>% 
  fit(mpg ~ ., data = mtcars) %>%
  multi_predict(new_data = mtcars[1:3, -1])
```

*However*, the API is still being developed. Currently, there is not an interface in the prediction functions to pass in the values of the parameters to make predictions with (`lambda`, in this case).

**What do I do about how my model handles factors or categorical data?**

Some modeling functions in R create indicator/dummy variables from categorical data when you use a model formula (typically using `model.matrix()`), and some do not. Some examples of models that do *not* create indicator variables include tree-based models, naive Bayes models, and multilevel or hierarchical models. The tidymodels ecosystem assumes a `model.matrix()`-like default encoding for categorical data used in a model formula, but you can change this encoding using `set_encoding()`. For example, you can set predictor encodings that say, "leave my data alone," and keep factors as is:

```{r}
#| label: "encodinginfo"
#| eval: false
set_encoding(
  model = "decision_tree",
  eng = "rpart",
  mode = "regression",
  options = list(
    predictor_indicators = "none",
    compute_intercept = FALSE,
    remove_intercept = FALSE
  )
)
```

::: callout-note
There are three options for `predictor_indicators`: - "none" (do not expand factor predictors) - "traditional" (apply the standard `model.matrix()` encoding) - "one_hot" (create the complete set including the baseline level for all factors)
:::

To learn more about encoding categorical predictors, check out [this blog post](https://www.tidyverse.org/blog/2020/07/parsnip-0-1-2/#predictor-encoding-consistency).

**What is the `defaults` slot and why do I need it?**

You might want to set defaults that can be overridden by the user. For example, for logistic regression with `glm`, it make sense to default `family = binomial`. However, if someone wants to use a different link function, they should be able to do that. For that model/engine definition, it has:

```{r}
#| label: "glm-alt"
#| eval: false
defaults = list(family = expr(binomial))
```

So that is the default:

```{r}
#| label: "glm-alt-show"
#| eval: false
logistic_reg() %>% translate(engine = "glm")

# but you can change it:

logistic_reg() %>%
  set_engine("glm", family = expr(binomial(link = "probit"))) %>% 
  translate()
```

That's what `defaults` are for.

Note that we wrapped `binomial` inside of `expr()`. If we didn't, it would substitute the results of executing `binomial()` inside of the expression (and that's a mess).

**What if I want more complex defaults?**

The `translate` function can be used to check values or set defaults once the model's mode is known. To do this, you can create a model-specific S3 method that first calls the general method (`translate.model_spec()`) and then makes modifications or conducts error traps.

For example, the ranger and randomForest package functions have arguments for calculating importance. One is a logical and the other is a string. Since this is likely to lead to a bunch of frustration and GitHub issues, we can put in a check:

```{r}
#| label: "rf-trans"
#| eval: false
# Simplified version
translate.rand_forest <- function (x, engine, ...){
  # Run the general method to get the real arguments in place
  x <- translate.default(x, engine, ...)
  
  # Check and see if they make sense for the engine and/or mode:
  if (x$engine == "ranger") {
    if (any(names(x$method$fit$args) == "importance")) 
      if (is.logical(x$method$fit$args$importance)) 
        rlang::abort("`importance` should be a character value. See ?ranger::ranger.")
  }
  x
}
```

As another example, `nnet::nnet()` has an option for the final layer to be linear (called `linout`). If `mode = "regression"`, that should probably be set to `TRUE`. You couldn't do this with the `args` (described above) since you need the function translated first.

**My model fit requires more than one function call. So....?**

The best course of action is to write wrapper so that it can be one call. This was the case with xgboost and keras.

**Why would I preprocess my data?**

There might be non-trivial transformations that the model prediction code requires (such as converting to a sparse matrix representation, etc.)

This would **not** include making dummy variables and `model.matrix` stuff. The parsnip infrastructure already does that for you.

**Why would I post-process my predictions?**

What comes back from some R functions may be somewhat... arcane or problematic. As an example, for xgboost, if you fit a multi-class boosted tree, you might expect the class probabilities to come back as a matrix (*narrator: they don't*). If you have four classes and make predictions on three samples, you get a vector of 12 probability values. You need to convert these to a rectangular data set.

Another example is the predict method for ranger, which encapsulates the actual predictions in a more complex object structure.

These are the types of problems that the post-processor will solve.

**Are there other modes?**

Not yet but there will be. For example, it might make sense to have a different mode when doing risk-based modeling via Cox regression models. That would enable different classes of objects and those might be needed since the types of models don't make direct predictions of the outcome.

If you have a suggestion, please add a [GitHub issue](https://github.com/tidymodels/parsnip/issues) to discuss it.

## Session information {#session-info}

```{r}
#| label: "si"
#| echo: false
small_session(pkgs)
```
